---
title: "Hair Fall Risk Factors — Exploratory Data Analysis"
author: "Vicente Just"
date: "`r Sys.Date()`"
output:
  html_document:
    toc: true
    toc_depth: 3
    toc_float: true
    theme: flatly
    highlight: tango
    df_print: paged
  github_document:
    toc: true
    toc_depth: 3
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(
  echo    = TRUE,
  message = FALSE,
  warning = FALSE,
  fig.align = "center"
)
```

## Overview

This analysis explores the **Predict Hair Fall** dataset, which collects data on
potential risk factors associated with hair loss. The goal is to understand the
structure of the data, explore relationships between variables, and evaluate
whether hair loss is driven by a single factor or by a combination of conditions.

---

## 1 — Dataset Loading and Initial Inspection

```{r load-libraries-and-data, echo = FALSE}
library(readr)
library(dplyr)
library(ggplot2)
library(datasets)
library(openxlsx)

# Load the dataset
HairFallH <- read_csv("Predict Hair Fall.csv")

# Preview the first records
head(HairFallH)
```

### Dataset Structure

```{r dataset-structure, echo = FALSE}
cat("Number of observations:", nrow(HairFallH), "\n")
cat("Number of variables:   ", ncol(HairFallH), "\n\n")

cat("Variable names:\n")
names(HairFallH)

cat("\nData types and structure:\n")
str(HairFallH)

cat("\nMissing values per variable:\n")
colSums(is.na(HairFallH))

# Remove rows with missing values (none found, but included as best practice)
HairFallH <- na.omit(HairFallH)

cat("\nDataset size after cleaning:", dim(HairFallH), "\n")

cat("\nStatistical summary:\n")
summary(HairFallH)
```

### Variable Overview

| Type | Variables |
|---|---|
| **Categorical** | Genetics, Hormonal Changes, Medical Conditions, Medications & Treatments, Nutritional Deficiencies, Stress, Poor Hair Care Habits, Environmental Factors, Smoking, Weight Loss |
| **Numerical** | Age, Hair Loss, Id |

**Potential analyses:**

- Frequency tables and bar charts for categorical variables
- Histograms and boxplots for numerical variables
- Chi-square tests for associations between categorical variables
- Correlation analysis between numerical variables
- t-test or ANOVA for group comparisons

---

## 2 — BOD Dataset Exploration

The built-in R dataset **BOD** (Biochemical Oxygen Demand) is used here to
demonstrate dataset inspection and export workflows.

### 2.1 — Exploratory Summary

```{r bod-exploration, echo = FALSE}
data(BOD)

cat("BOD DATASET — BASIC INFORMATION\n")
cat("Observations:", nrow(BOD), "\n")
cat("Variables:   ", ncol(BOD), "\n")
cat("Dimensions:  ", dim(BOD), "\n\n")

cat("Variable names:\n")
names(BOD)

cat("\nData structure:\n")
str(BOD)

cat("\nMissing values:\n")
colSums(is.na(BOD))

cat("\nStatistical summary:\n")
summary(BOD)

cat("\nDuplicated rows:\n")
duplicated(BOD)
```

The BOD dataset contains 6 observations and 2 numeric variables:

- **Time**: days elapsed
- **demand**: biochemical oxygen demand (mg/L)

No missing values or duplicates are present. Given its small size and temporal
structure, this dataset is well-suited for scatter/line plots, linear or
non-linear regression fitting, and organic matter degradation rate analysis.

### 2.2 — Exporting to Excel

```{r bod-export, echo = FALSE}
file_path <- file.path(getwd(), "BOD_dataset.xlsx")
write.xlsx(BOD, file_path)

cat("File exported to:", file_path, "\n")
if (file.exists(file_path)) {
  cat("✓ Export successful\n")
} else {
  cat("✗ Export failed\n")
}
```

---

## 3 — Stress, Nutritional Deficiencies, and Hair Loss

### 3.1 — Stress Level by Nutritional Deficiency

```{r stress-nutritional, echo = FALSE}
library(dplyr)

# Cross-tabulation: Stress vs Nutritional Deficiencies
stress_nd_table <- table(HairFallH$Stress, HairFallH$`Nutritional Deficiencies`)
prop_stress_nd   <- prop.table(stress_nd_table)
round(prop_stress_nd, 2)
```

Key observations:

- **Low stress**: highest deficiencies in Biotin, Omega-3, and Vitamin A (≈ 0.04 each)
- **Moderate stress**: Protein, Selenium, and Vitamin D are most prevalent
- **High stress**: Vitamin D deficiency stands out
- **Zinc** shows a remarkably consistent pattern across all stress levels, suggesting its influence is independent of stress

The most variable nutrient is **Selenium**, which peaks at 0.45 under high stress
versus ~0.02 in moderate/low conditions — indicating a strong interaction with
stress.

### 3.2 — Stress Level and Hair Loss

```{r stress-hairloss, echo = FALSE}
# Contingency table
stress_hl_table <- table(HairFallH$`Hair Loss`, HairFallH$Stress)
cat("Contingency table — Hair Loss vs Stress:\n")
print(stress_hl_table)

prop_stress_hl <- prop.table(stress_hl_table)
cat("\nProportions:\n")
round(prop_stress_hl, 2)

# Bar chart
barplot(
  prop_stress_hl,
  beside  = TRUE,
  col     = c("#262626", "#b287fd"),
  legend  = c("No Hair Loss", "Hair Loss"),
  main    = "Hair loss proportion by stress level",
  xlab    = "Stress level",
  ylab    = "Proportion"
)
```

Proportions are broadly similar across stress levels, though individuals with
**moderate stress** show a slightly higher incidence of hair loss. Stress appears
to be a contributing — rather than determining — factor.

### 3.3 — Nutritional Deficiency and Hair Loss Count

```{r nutritional-hairloss, echo = FALSE}
# Count per nutritional deficiency
deficiency_count <- table(HairFallH$`Nutritional Deficiencies`)
cat("Number of individuals per nutritional deficiency:\n")
print(deficiency_count)

# Proportions per deficiency
nd_hl_table  <- table(HairFallH$`Nutritional Deficiencies`, HairFallH$`Hair Loss`)
proportions  <- prop.table(nd_hl_table, margin = 1)
cat("\nHair loss proportion by nutritional deficiency:\n")
print(round(proportions, 3))

# Focus on Hair Loss = 1
prop_hairloss <- proportions[, "1"]
cat("\nProportion with hair loss per deficiency:\n")
print(round(prop_hairloss, 3))
```

All deficiency types yield proportions between 0.45 and 0.55, with no single
deficiency exceeding ~55% incidence. This strongly suggests that hair loss is
**multifactorial** rather than tied to one specific nutritional deficit.

Highest proportions: **Magnesium** (0.548), **Protein** (0.522), **Iron** (0.513)  
Lowest proportions: **Omega-3** (0.457), **Vitamin E** (0.458), **Biotin** (0.465)

### 3.4 — Smokers Under the Age Mean with Hair Loss, by Stress Level

```{r smokers-filtered, echo = FALSE}
age_mean <- mean(HairFallH$Age)
cat("Age mean:", age_mean, "\n")

filtered_34 <- subset(
  HairFallH,
  Smoking == "Yes" &
    Age < age_mean &
    HairFallH$`Hair Loss` == 1
)

cat("\nCount by stress level:\n")
table(filtered_34$Stress)
```

### 3.5 — Youngest Patients: High Stress, Smoking, Hair Loss, and Medication

```{r youngest-patients, echo = FALSE}
min_age <- min(HairFallH$Age)

youngest <- subset(
  HairFallH,
  Age == min_age &
    HairFallH$`Hair Loss` == "1" &
    Smoking == "Yes" &
    Stress == "High" &
    HairFallH$`Medications & Treatments` != "No Data"
)

cat("Nutritional deficiencies in this subgroup:\n")
table(youngest$`Nutritional Deficiencies`)
```

Three patients match this profile, each with a **different** nutritional
deficiency — further confirming the multifactorial nature of hair loss observed
in section 3.3.

### 3.6 — Hair Loss by Age Group and Stress Level

```{r age-group-stress, echo = FALSE}
# Define age groups
AgeGroup <- cut(
  HairFallH$Age,
  breaks = c(0, 20, 30, 40, 50),
  labels = c("≤20", "21–30", "31–40", "41–50")
)

# Filter to individuals with hair loss
hair_loss_data <- HairFallH[HairFallH$`Hair Loss` == "1", ]
age_group_hl   <- AgeGroup[HairFallH$`Hair Loss` == "1"]

# Count by age group
cat("Hair loss count by age group:\n")
age_hl_table <- table(age_group_hl)
print(age_hl_table)

barplot(
  age_hl_table,
  col  = c("#262626", "#b287fd", "#b2f142", "#d8d9dd"),
  main = "Hair loss by age group",
  xlab = "Age range",
  ylab = "Number of individuals"
)

# Age group vs stress level
cat("\nAge group vs stress level (hair loss individuals):\n")
age_stress_table <- table(AgeGroup, HairFallH$Stress)
print(age_stress_table)

barplot(
  age_stress_table,
  beside = TRUE,
  col    = c("#262626", "#b287fd", "#b2f142", "#d8d9dd"),
  main   = "Stress level by age group (hair loss individuals)",
  xlab   = "Stress level",
  ylab   = "Number of individuals"
)

legend(
  "topleft",
  legend = rownames(age_stress_table),
  fill   = c("#262626", "#b287fd", "#b2f142", "#d8d9dd"),
  title  = "Age group"
)
```

All three stress levels are meaningfully represented across every age group,
confirming that neither age nor stress alone drives hair loss — the condition
spans the full range of both dimensions.

---

## 4 — Descriptive Statistics and Advanced Visualisations

### 4.1 — Descriptive Statistics

```{r descriptive-stats, echo = FALSE}
cat("DATASET STRUCTURE\n")
str(HairFallH)

cat("\nFull summary:\n")
summary(HairFallH)

# Numerical variables
cat("\nNumerical variables:\n")
summary(HairFallH[, c("Id", "Age")])

# Categorical variables
categorical_vars <- c(
  "Genetics", "Hormonal Changes", "Medical Conditions",
  "Stress", "Poor Hair Care Habits", "Smoking",
  "Weight Loss", "Hair Loss"
)

for (var in categorical_vars) {
  cat("\n====", var, "====\n")
  print(table(HairFallH[[var]]))
  cat("Proportions (%):",
      round(prop.table(table(HairFallH[[var]])) * 100, 1), "\n")
}
```

**Variable roles:**

- **Age**: the only continuous variable; shows mean, median, range and distribution
- **Id**: row identifier — no analytical value
- **Hair Loss**: binary outcome variable (0 = no loss, 1 = loss)
- **Stress**: ordinal with 3 levels (Low / Moderate / High)
- **Medical Conditions, Genetics, Hormonal Changes…**: categorical risk factors

### 4.2 — Age Distribution by Hair Loss (Boxplot)

```{r boxplot-age-hairloss, echo = FALSE}
boxplot(
  Age ~ HairFallH$`Hair Loss`,
  data  = HairFallH,
  main  = "Age distribution by hair loss status",
  xlab  = "Hair loss (0 = No, 1 = Yes)",
  ylab  = "Age",
  col   = c("#262626", "#b287fd")
)
```

The age distributions for both groups are nearly identical, indicating that
**age alone is not a determining factor** for hair loss. Its influence is likely
amplified by other variables such as genetics, stress, medical conditions, and
nutritional deficiencies — consistent with a multifactorial model.

### 4.3 — Medical Conditions and Hair Loss (Heat Map)

```{r heatmap-medical, echo = FALSE}
# Contingency table: Medical Conditions vs Hair Loss
med_hl_table  <- table(HairFallH$`Medical Conditions`, HairFallH$`Hair Loss`)
prop_med_hl   <- prop.table(med_hl_table, margin = 1)
prop_med_hl   <- round(prop_med_hl, 2)

cat("Medical conditions vs hair loss (row proportions):\n")
prop_med_hl

# Convert to data frame for ggplot2
df_med_hl <- as.data.frame(prop_med_hl)
colnames(df_med_hl) <- c("MedicalCondition", "HairLoss", "Proportion")

ggplot(df_med_hl, aes(x = HairLoss, y = MedicalCondition, fill = Proportion)) +
  geom_tile(color = "white") +
  scale_fill_gradient(low = "#262626", high = "#b287fd") +
  labs(
    title = "Heat map: Medical conditions vs Hair loss",
    x     = "Hair loss",
    y     = "Medical condition",
    fill  = "Proportion"
  ) +
  theme_minimal()
```

Among individuals **with hair loss (1)**, the most relevant medical conditions are:

- **Androgenetic Alopecia** and **Alopecia Areata** — expected, as they are
  defined by hair loss
- **Seborrheic Dermatitis** — consistent with known weakening of the hair shaft
  through excess sebum production

### 4.4 — Age Distribution of Hair Loss Individuals: Normality Assessment

```{r histogram-normality, echo = FALSE}
# Filter to individuals with hair loss
hl_subset <- subset(HairFallH, HairFallH$`Hair Loss` == "1")

cat("Sample size:", nrow(hl_subset), "\n")

age_mean <- mean(hl_subset$Age)
age_sd   <- sd(hl_subset$Age)
cat("Mean age:", age_mean, "\n")
cat("Std dev: ", age_sd, "\n")

fd_breaks <- nclass.FD(hl_subset$Age)

hist(
  hl_subset$Age,
  breaks = fd_breaks,
  col    = "#b287fd",
  prob   = TRUE,
  main   = "Age distribution — individuals with hair loss",
  xlab   = "Age",
  ylab   = "Density",
  ylim   = c(0, 0.05)
)

legend(
  "topright",
  legend = c("Empirical density", "Theoretical normal"),
  col    = c("red", "#262626"),
  lwd    = 2,
  lty    = c(1, 2)
)

# Empirical density curve
lines(density(hl_subset$Age), col = "red", lwd = 2)

# Theoretical normal curve
curve(
  dnorm(x, mean = age_mean, sd = age_sd),
  add = TRUE, col = "black", lwd = 2, lty = 2
)
```

The **empirical density (red)** and **theoretical normal (black dashed)** curves
diverge clearly. The distribution is right-skewed, with a pronounced peak around
20–25 years and a sharp decline after 40.

**Conclusion:** The age variable among hair loss individuals does **not** follow
a normal distribution. This means subsequent statistical comparisons should use
**non-parametric tests** rather than t-tests or ANOVA.